A Mechanised Library for Generic (Co)Programming

### 2024-02-28 paper/paper.tex
Extractable Containers

Here consider instead representing the positions of a
container using a validity predicate which assigns shapes to positions.
A container extension is then defined as follows:
\begin{coqcode}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{coqcode}
This approach solves the problem of the unsafe casts, since the code extracted
for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
OCaml equivalent to \coq{Pos}.

Additionaly we would like to reason about equality of container extensions. The
contents of the container extensions need to be accessed via a
function \coq{contents} whose domain depends on the shape of the container.
This means that for a given type of shapes and positions, the equality of two
objects of a container extension will need to deal with heterogeneous equality.
Since we want to remain axiom-free, one way to deal with this is to rely on a
\emph{decidable} validity predicate.

Finally we can define the type of containers as follows:
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{coqcode}
where \coq{valid} is the First, containers are defined as pairs of shape and
position types that have a \emph{decidable} validity predicate. We represent
decidable predicates as functions that return a boolean and lift booleans to
propositions similarly to the implementation provided by
SSReflect~\cite{GonthierL09}.  For simplicity, we define the dependent pair of
valid positions as a record, which will be extracted to a singleton type with
the validity proof erased by Coq's code extraction mechanism.

Then, we define \emph{container extensions} as follows:
\begin{coqcode}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{coqcode}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{coqcode}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}

>>>>

Here consider instead representing the positions of a container using a validity
predicate \coq{Valid : Shape -> Pos -> bool} which assigns shapes to positions.
Moreover, it is crucial for our purposes of remaining axiom-free that this
predicate is \emph{decidable} in order to be able to reason about equality of
container extensions and, as a result, the contents of the container extensions
need to be accessed via a function \coq{contents} whose domain depends on the
shape of the container.  This means that for a given type of shapes and
positions, the equality of two objects of a container extension will need to
deal with heterogeneous equality.
\mpav{This part needs more explanation.}

\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) :=
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{coqcode}
where \coq{valid} is the \emph{decidable} validity predicate. Note that booleans
can be lifted to propositions similarly to the implementation provided by
SSReflect~\cite{GonthierL09}. This allows us to define the dependent pair of
valid positions as a record which will be extracted to a singleton type since
the validity proof is erased by Coq's code extraction mechanism.

We now define \emph{container extensions} as follows:
\begin{coqcode}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{coqcode}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{coqcode}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}

% A container extension is then defined as follows:
% \begin{coqcode}
% Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
% Record App (X : Type) := { shape    : Shape;
%                           contents : {p | Valid shape p} -> X }.
% \end{coqcode}
% This approach solves the problem of the unsafe casts, since the code extracted
% for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
% OCaml equivalent to \coq{Pos}.

Finally we can define the type of containers as follows:
